---
title: "Exercises"
output:
  rmarkdown::html_document:
    code_folding: 'hide'
    toc: true
    toc_float: true
    smart: true
    number_sections: true
    highlight: 'textmate'
    self_contained: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,results = "markup",collapse = TRUE)
library(tidyverse)
library(magrittr)
library(lubridate)
library(pryr)
```

# Vectors

## Considering the vectors **x** and **y**, perform the following operations:
```{r}
x <- 8
y <- 2
```

### Add *x* and *y*
```{r}
x+y
```
### Subtract *x* and *y*
```{r}
x-y
```
### Multiply *x* by *y*
```{r}
x*y
```
### Remainder of dividion of *x* by *y*
```{r}
x %% y
```

## Now suppose that you have the following vector:
  + x <- 1:999
  
### Which are even numbers?
```{r, results='hide'}
x <- 1:999
even_numbers <- x %% 2 == 0
even_numbers
```

### How about odd ones?
```{r,results='hide'}
odd_numbers <- x %% 2 != 0
odd_numbers
```

## Using the *iris* data set, perform the following

### Extract *Species* variable and assign it to a different
```{r}
iris_species <- iris$Species
```

### What's the type of vector?
```{r}
str(iris_species)

# alternative
is.factor(iris_species)

# alternative
type_sum(iris_species)
```

### How many observations are there?
```{r}
length(iris_species)
```


### What are the levels?
```{r}
levels(iris_species)
```

# Lists

## Exercise 1

### LCreate a cars list
Using the *mtcars* dataset, extract the number of cylinders and horsepower, and car names and put them in a list named cars
```{r}
# Have a look at the data
head(mtcars)

# cyl is the cylinders, horsepower is hp and names is rows
cylinders <- mtcars$cyl
horsepower <- mtcars$hp
carnames <- rownames(mtcars)

#Put in a list
cars <- list(cylinders=cylinders, HP=horsepower, Car_Names=carnames)
```

### Do the same for the *rock* dataset, extracting area and shape
```{r}
head(rock)

rock_area <- rock$area
rock_shape <- rock$shape

rock_list <- list(Area=rock_area, Shape=rock_shape)
```

## Exercise 2

### Create a list which contains both lists created before
```{r}
rock_cars <- list(Cars=cars, Rocks=rock_list)
```

### Visualise the list
```{r}
str(rock_cars)
```

### How do you get the car names?
```{r}
rock_cars$Cars$Car_Names

# alternative
rock_cars[['Cars']][['Car_Names']]

# alternative
rock_cars[[1]][[3]]
```

# Strings

  
## Exercise 1
  For this exercise, we're going to be using the *fruit* data from the stringr package

### Get the fruits which begin with b
```{r}
b_fruits <- str_detect(fruit, '^b')

fruit[b_fruits]
```

### What about fruits that begin with vowels?
```{r}
vowel_fruit <- str_detect(fruit,'^[aeiou]')

fruit[vowel_fruit]
```

## Exercise 2
  For this exercise, we're using the *senteces* data 

### How many sentences are there?
```{r}
length(sentences)
```

### How do you find the length of a sentence?
```{r}
str_length(sentences[1])
```

### How long is the longest sentence?
```{r}
max( str_length(sentences) )
```

### What's that sentence?
```{r}
#Find maximum value for all sentence lengths
max(str_length(sentences))

# Find sentence with that length
long_sentence <- str_length(sentences)==57

#Getting the sentence
sentences[long_sentence]
```

### Reorder the sentences to alphabetical order
```{r, results='hide'}
# Order strings according to alphabetical order, remembering to set language, although not necessary in this case
alphabetical <- str_order(sentences,locale = 'en')

# reorder according to alphabetical order
sentences[alphabetical]
```

# Factors

  For these exercises, we're going to use the *forcats::gss_cat* dataset. It's a sample dataset from the [General Social Survey](http://gss.norc.org), i.e. a US-based survey run by NORC, an independent organisation at the University of Chicago
  
## Exercise 1
  Have a look at the dataset
  
### Get the levels of any of the variables
```{r}
# Using race variable
levels(gss_cat$race)

# Alternative
gss_cat %>% 
  count(race)
```

### What is the most common religion in the survey?
```{r}
gss_cat %>% 
  count(relig)

# alternative
summary(gss_cat$relig)
```


# Data Transformation

## Exercise 1

  Using the *nycflights13::flights* dataset, do the following:

### Find flights which had an arrival delay of two or more hours
```{r}
nycflights13::flights %>% 
  filter(arr_delay>=2)
```

### Find flights departed in summer (July, August, September)
```{r}
nycflights13::flights %>% 
  filter(month==7| month==8| month==9)
```

### Try out the **between()** function to simplify the previous code
```{r}
nycflights13::flights %>% 
  filter(between(month,7,9))
```

## Exercise 2

### Sort flights to find those with greatest departure delay, putting earliest first
```{r}
nycflights13::flights %>% 
  arrange(dep_delay)
```

### Which flights travelled the longest?
```{r}
nycflights13::flights %>% 
  arrange(desc(air_time))
```

### Which travelled the shortest?
```{r}
nycflights13::flights %>% 
  arrange(air_time)
```

## Exercise 3

### Propose different ways to select *dep_time*, *dep_delay*, *arr_time*, *arr_delay* from flights
```{r}
nycflights13::flights %>% 
  select(dep_time,dep_delay,arr_time,arr_delay)

# Alternative
nycflights13::flights %>% 
  select(starts_with('dep'),starts_with('arr'))
```

### What is wrong with the following code?
>select(flights,contains('TIME'))

